Assessing primitives performance\\ on multi-stage execution

### 2018-05-18 paper.tex
fix typos

Most of the Cog VM code base is written in Slang, a subset of Smalltalk. Slang is compiled to C and then to native code through standard C compilers. The execution engine (the memory manager, the interpreter and the baseline JIT) are entirely written in Slang. The two main purposes of using Slang code over plain C are:
\begin{itemize}
	\item to specify with annotations what function needs to be inlined or duplicated with constant parameters, and
	\item to be able to simulate VM execution, executing the Slang code as Smalltalk code on top of a compiled VM for debugging purposes.
\end{itemize}

@@

To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end. Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code. This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance. Primitives redefined in such way can be entirely or partially redefined. In the later case, only the common cases are re-implemented on top of the JIT back-end. If an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.

@@

\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters. The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters). If no order is specified, the VM assumes the byte ordering (ASCII order) is the order to compare against, and compares directly the bytes of the ByteStrings. If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compare the bytes through the indirection order. Having the order optional allows to have quicker string comparison in the common case where the ASCII order is used. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

>>>>

Most of the Cog VM code base is written in Slang, a subset of Smalltalk. Slang is compiled to C and then to native code through standard C compilers. The execution engine (the memory manager, the interpreter and the baseline JIT) is entirely written in Slang. The two main purposes of using Slang code over plain C are:
\begin{itemize}
	\item to specify with annotations what function needs to be inlined or duplicated with constant parameters, and
	\item to be able to simulate VM execution, executing the Slang code as Smalltalk code on top of a compiled VM for debugging purposes.
\end{itemize}

@@

To solve this performance bottleneck, the Cog VM allows one to redefine primitives in Cog's Register Transfer Language (RTL), an abstract assembly which is compiled to native code through Cog's JIT back-end. Such primitives are then generated to native code at runtime when a method annotated with the primitive is compiled to native code. This allows the primitive to be performed in the client stack with the client calling convention, saving register moves and the stack switch dance. Primitives redefined in such way can be entirely or partially redefined. In the latter case, only the common cases are re-implemented on top of the JIT back-end. If an unimplemented case happens at runtime, the generated native code calls the C primitive instead, adding overhead only in uncommon cases.

@@

\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters. The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters). If no order is specified, the VM assumes the byte ordering (ASCII order) is the order to compare against, and compares directly the bytes of the ByteStrings. If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order. Having the order optional allows to have quicker string comparison in the common case where the ASCII order is used. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.
